# Introduction

R is a open-source program that is commonly used in Statistics. It runs on almost every platform and is completely free and is available at [www.r-project.org](www.r-project.org). Most of the cutting-edge statistical research is first available on R. 

R is a script based language, so there is no point and click interface. (Actually there are packages that attempt to provide a point and click interface, but they are still somewhat primitive.) While the initial learning curve will be steeper, understanding how to write scripts will be valuable because it leaves a clear description of what steps you performed in your data analysis. Typically you will want to write a script in a separate file and then run individual lines. The saves you from having to retype a bunch of commands and speeds up the debugging process.

This document is a very brief introduction to using R in my course. I highly recommend downloading and reading/skimming the manual “An Introduction to R” which is located at [cran.r-project.org/doc/manuals/R-intro.pdf](http://cran.r-project.org/doc/manuals/R-intro.pdf).

Finding help about a certain function is very easy. At the prompt, just type `help(function.name)` or `?function.name`. If you don't know the name of the function, your best bet is to go the the web page www.rseek.org which will search various R resources for your keyword(s). Another great resource is the coding question and answer site [stackoverflow](http://stackoverflow.com).

The basic editor that comes with R works fairly well, but you should consider running R through the program RStudio which is located at [rstudio.com](http://www.rstudio.org). This is a completely free Integrated Developement Environment that works on Macs, Windows and a couple of flavors of Linux. It simplifies a bunch of more annoying aspects of the standard R GUI and supports things like tab completion.


When you first open up R (or RStudio) the console window gives you some information about the version of R you are running and then it gives the prompt `>`.  This prompt is waiting for you to input a command. The prompt + tells you that the current command is spanning multiple lines. In a script file you might have typed something like this:

    for( i in 1:5 ){
        print(i)
    }

But when you copy and paste it into the console in R you'll see something like this:

    > for (i in 1:5){
    +    print(i)  
    + }

If you type your commands into a file, you won't type the `>` or `+` prompts. For the rest of the tutorial, I will show the code as you would type it into a script and I will show the output being shown with two hashtags (`##`) before it to designate that it is output.


## R as a simple calculator

Assuming that you have started R on whatever platform you like, you can use R as a simple calculator. At the prompt, type 2+3 and hit enter. What you should see is the following
```{r}
# Some simple addition
2+3
```

In this fashion you can use R as a very capable calculator.

```{r}
6*8
4^3
exp(1)   # exp() is the exponential function
```

R has most constants and common mathematical functions you could ever want. `sin()`, `cos()`, and other trigonometry functions are available, as are the exponential and log functions `exp()`, `log()`. The absolute value is given by `abs()`, and `round()` will round a value to the nearest integer. 

```{r}
pi     # the constant 3.14159265...
sin(0)
log(5) # unless you specify the base, R will assume base e
log(5, base=10)  # base 10
```

Whenever I call a function, there will be some arguments that are mandatory, and some that are optional and the arguments are separated by a comma. In the above statements the function `log()` requires at least one argument, and that is the number(s) to take the log of. However, the base argument is optional. If you do not specify what base to use, R will use a default value. You can see that R will default to using base $e$ by looking at the help page (by typing `help(log)` or `?log` at the command prompt).

Arguments can be specified via the order in which they are passed or by naming the arguments. So for the `log()` function which has arguments `log(x, base=exp(1))`. If I specify which arguments are which using the named values, then order doesn't matter.

```{r}
# Demonstrating order does not matter if you specify
# which argument is which
log(x=5, base=10)   
log(base=10, x=5)
```

But if we don't specify which argument is which, R will decide that `x` is the first argument, and `base` is the second.

```{r}
# If not specified, R will assume the second value is the base...
log(5, 10)
log(10, 5)
```

When I specify the arguments, I have been using the `name=value` notation and a student might be tempted to use the `<-` notation hereSee next section.. Don't do that as the `name=value` notation is making an association mapping and not a permanent assignment.

## Assignment

We need to be able to assign a value to a variable to be able to use it later. R does this by using an arrow `<-` or an equal sign `=`. While R supports either, for readability, I suggest people pick one assignment operator and stick with it. I personally prefer to use the arrow. Variable names cannot start with a number, may not include spaces, and are case sensitive.

```{r}
tau <- 2*pi       # create two variables
my.test.var = 5   # notice they show up in 'Environment' tab in RStudio!
tau
my.test.var
tau * my.test.var
```

As your analysis gets more complicated, you'll want to save the results to a variable so that you can access the results later^[To paraphrase Beyonce, “Cause if you liked it, then you should have put a name on it.”]. *If you don't assign the result to a variable, you have no way of accessing the result.* ^[This isn't strictly true, the variable `.Last.value` always has the result of the last expression evaluated, but you can't go any farther back.]

## Scripts and RMarkdown

One of the worst things about a pocket calculator is there is no good way to go several steps and easily see what you did or fix a mistake (there is nothing more annoying than re-typing something because of a typo. To avoid these issues I always work with script (or RMarkdown) files instead of typing directly into the console. You will quickly learn that it is impossible to write R code correctly the first time and you'll save yourself a huge amount of work by just embracing scripts (and RMarkdown) from the beginning. Furthermore, having a script file fully documents how you did your analysis, which can help when writing the methods section of a paper. Finally, having a script makes it easy to re-run an analysis after a change in the data (additional data values, transformed data, or removal of outliers).

It often makes your script more readable if you break a single command up into multiple lines. R will disregard all whitespace (including line breaks) so you can safely spread your command over as multiple lines. Finally, it is useful to leave comments in the script for things such as explaining a tricky step, who wrote the code and when, or why you chose a particular name for a variable. The `#` sign will denote that the rest of the line is a comment and R will ignore it.

### R Scripts (.R files)

The first type of file that we'll discuss is a traditional script file. To create a new .R script in RStudio go to `File -> New File -> R Script`. This opens a new window in RStudio where you can type commands and functions as a common text editor. Type whatever you like in the script window and then you can execute the code line by line (using the run button or its keyboard shortcut to run the highlighted region or whatever line the curser is on) or the entire script (using the source button). Other options for what piece of code to run are available under the Code dropdown box.

An R script for a homework assignment might look something like this:

```{r, eval=FALSE}
# Problem 1 
# Calculate the log of a couple of values and make a plot
# of the log function from 0 to 3
log(0)
log(1)
log(2)
x <- seq(.1,3, length=1000)
plot(x, log(x))

# Problem 2
# Calculate the exponential function of a couple of values
# and make a plot of the function from -2 to 2
exp(-2)
exp(0)
exp(2)
x <- seq(-2, 2, length=1000)
plot(x, exp(x))
```
This looks perfectly acceptable as a way of documenting what you did, but this script file doesn't contain the actual results of commands I ran, nor does it show you the plots. Also anytime I want to comment on some output, it needs to be offset with the commenting character `#`. It would be nice to have both the commands and the results merged into one document. This is what the R Markdown file does for us.

### R Markdown (.Rmd files)

When I was a graduate student, I had to tediously copy and past tables of output from the R console and figures I had made into my Microsoft Word document. Far too often I would realize I had made a small mistake in part (b) of a problem and would have to go back, correct my mistake, and then redo all the laborious copying. I often wished that I could write both the code for my statistical analysis and the long discussion about the interpretation all in the same document so that I could just re-run the analysis with a click of a button and all the tables and figures would be updated by magic. Fortunately that magic^[Clark's third law states “Any sufficiently advanced technology is indistinguishable from magic.”] now exists.

To create a new R Markdown document, we use the `File -> New File -> R Markdown...` dropdown option and a menu will appear asking you for the document title, author, and preferred output type. In order to create a PDF, you'll need to have LaTeX installed, but the HTML output nearly always works and I've had good luck with the MS Word output as well.

The R Markdown is an implementation of the Markdown syntax that makes it extremely easy to write webpages and give instructions for how to do typesetting sorts of things. This syntax was extended to allow use to embed R commands directly into the document. Perhaps the easiest way to understand the syntax is to look at an at the [RMarkdown website](http://rmarkdown.rstudio.com).

The R code in my document is nicely separated from my regular text using the three backticks and an instruction that it is R code that needs to be evaluated. The output of this document looks good as a HTML, PDF, or MS Word document. I have actually created this entire document using RMarkdown.


## Exercises

Create an RMarkdown file that solves the following exercises.

1. Calculate $\log\left(6.2\right)$ first using base $e$ and second using base $10$. To figure out how to do different bases, it might be helpful to look at the help page for the `log` function.

2. Calculate the square root of 2 and save the result as the variable named sqrt2. Have R display the decimal value of sqrt2. *Hint: use Google to find the square root function. Perhaps search on the keywords  "R square root function".*
